Approximating the matrix exponential derivative of the CTMC generator matrix, with application to global SARS-CoV-2

### 2023-03-18 manuscript/main.tex
hyphen when adjective

 Let data $\y$ be the $N$-vector with elements $y_n\in \{1,\dots,S\}$, the observed states of each observed biological specimen $n \in \{1,\dots,N\}$.  At any node $v$, the likelihood takes the form \citep{ji2020gradients}
 \begin{align}
p(\y|\ttheta) = \p_v^T \q_v \, ,
 \end{align}
where $\p_v$ and $\q_v$ are the post-order and pre-order partial likelihood vectors, respectively.  The former describes the probability of the observed states for all observed specimens (i.e., leaf nodes) that descend from node $v$, conditioned on the state at node $v$. The latter describes analogous probabilities for all observed specimens not descending from node $v$.  If we suppose that node $u$ gives rise to two child nodes, $v$ and $w$, then 
\begin{align}\label{eq:recursions}
	\p_u = \P_v \p_v \circ \P_w \p_w\, ,  \quad \q_v= \P_v^T \left( \q_u \circ  \P_w \p_w\right) \, .
\end{align}
Using the chain rule and the fact that $\p_v$ does not depend on $\P_v$, one may write the likelihood's derivative with respect to a single parameter $\theta_d$ thus:
\begin{align}\label{eq:phyloGrad}
	\frac{\partial}{\partial \theta_d}  p(\y|\ttheta) 
	&=   \sum_{v=1}^{2N-2} \mbox{tr} \left( \frac{\partial (\p_v^T\q_v)}{\partial \P_v} \frac{\partial \P^T_v}{\partial \theta_d} \right)
	= \sum_{v=1}^{2N-2} \mbox{tr} \left(  (\q_u\circ \P_w\p_w) \p_v^T\bigg( \frac{\partial e^{\gamma t_v \QQ}}{\partial \theta_d}\bigg)^{\!T}\right)  \\ \nonumber
	&=\sum_{v=1}^{2N-2} \p_v^T \Biggl(\sum_{s=1}^S \sum_{s'=1}^S \frac{\partial e^{\gamma t_v \QQ}}{\partial q_{ss'}} \frac{\partial q_{ss'}}{\partial \theta_d} \Biggr)^{\!T} (\q_u\circ \P_w\p_w)  \, .
\end{align}
Whereas the recursions of \eqref{eq:recursions} facilitate fast likelihood computation, inferring $\ttheta$ using gradient-based approaches such as HMC (Section \ref{sec:surrHmc}) requires a large number of repeated evaluations of the matrix exponential derivative.  These computations become particularly onerous when one opts for a gold-standard mixed effect model \textcolor{red}{(MAGEE)} and specifies
\begin{align}\label{eq:mixed}
\log q_{ss'} = b_{ss'} + \epsilon_{ss'} \,, \quad s\neq s'\, .
\end{align} 
Here, the fixed-effects $b_{ss'}$ are elements of some non-random matrix $\B$, and the random effects $\epsilon_{ss'}$ are mutually independent \emph{a priori} and inferred as model parameters.  The dimension $D$ of $\ttheta$ in this model is $\mathcal{O}(S^2)$, so the usual $\mathcal{O}(DNS^5)$ cost of the matrix expoential derivative becomes a massive $\mathcal{O}(NS^7)$.  In this context, \textcolor{red}{MAGEE} show that the first-approximation \eqref{eq:firstOrder} achieves a considerable speedup over the exact derivative, while facilitating high-quality proposals in the context of surrogate HMC.  In the following section, we use this method to analyze the global spread of SARS-CoV-2 and show that the first-order approximation maintains its performance for an even higher-dimensional problem than previously considered.

@@

\begin{figure}[!t]
	\centering
	\includegraphics[width=\linewidth]{fixedEffects.pdf}
	\vspace{-2em}
	\caption{Matrix predictors described in \eqref{eq:linear} combine in a linear manner to form the fixed effect matrix $\B$ featured within the mixed effects regression model \eqref{eq:mixed}.  Air traffic proximities \emph{(\textbf{a})} are proportional the number of air passengers exchanged between airports within respective regions \citep{holbrook2021massive}.  Intracontinental proximities \emph{(\textbf{b})} take values between -1 for regions on different continents to 1 for adjacent regions.  The Hubei asymmetry \emph{(\textbf{c})} roughly characterizes the Hubei quarantine of early 2020.}\label{fig:FE}
\end{figure}

We use the phylogenetic CTMC framework to model the early spread of SARS-CoV-2---the virus responsible for the ongoing COVID-19 pandemic---based on 285 observed viral samples collected from $S=31$ regions worldwide between December 24, 2019 and March 19, 2020. These regions comprise 13 provinces within China and 18 countries without.  Understanding the manner in which viruses travel between human populations is an object of ongoing study, and phylogeographic analyses point to the central role of travel networks including those measured by airline passenger counts \citep{holbrook2021massive} or google mobility data \citep{worobey2020emergence}.  Here, we include three such predictors of travel in our CTMC model by expanding the fixed effects in regression model \eqref{eq:mixed} to take the form
\begin{align}\label{eq:linear}
	\B = \theta_1 \X_1 + \theta_2  \X_2 + \theta_3 \X_3 \, ,
\end{align} 
for each $\X$ a fixed $44\times 44$ matrix predictor and each $\theta$ a real-valued regression coefficient, where we have expanded the number of regions between which viruses may travel by including two additional Chinese provinces and 11 additional countries.  Figure \ref{fig:FE} presents the three predictors of interest: $\X_1$ contains air travel proximities between locations as measured by annualized air passenger counts between airports contained within a region \citep{holbrook2021massive};  $\X_2$ contains intracontinental proximities arising from physical distances when two regions inhabit the same continent and fixed at the minimum otherwise; finally, $\X_3$ describes the Hubei asymmetry, i.e., the non-existence of human travel out of the Hubei province in early 2020.

@@

We generate 8 million MCMC samples in this manner, saving 1 in 10,000 Markov chain states, in order to guarantee a minimum effective sample size (ESS) greater than 100.  By far, the worst-mixing parameter  is the global scale $\tau$, which obtains an ESS of 185.  The three fixed-effects regression coefficients $\theta_1$, $\theta_2$ and $\theta_3$ obtain ESS's of 721, 721 and 644, respectively.  The median and minimum ESS for the 1,892 random effects $\epsilon_{ss'}$ are 721 and 190, respectively.  We note that, after thinning and removing burn-in, the sample only consists of 721 states, so ESS of 721 implies negligible autocorrelation between samples.  The left plot of Figure \ref{fig:post} presents posterior densities for the fixed effect coefficients from regression \eqref{eq:mixed}.  The air traffic proximity, intracontinental proximity and Hubei asymmetry coefficients have posterior means and 95\% credible intervals of 0.76 (0.50, 1.03), 0.04 (-0.04, 0.12) and 0.27 (-0.39, 0.78), respectively. While these results suggest a positive association between each predictor matrix and the infinitesimal generator matrix $\QQ$, the only predictor with a statistically significant association is air traffic proximity.  This result agrees with a previous phylogeographic analysis of the global spread of influenza \citep{holbrook2021massive}.  The right plot of Figure \ref{fig:post} presents the posterior mean for generator $\QQ$.  As one may expect, the matrix looks similar to that of the air traffic predictor matrix in Figure \ref{fig:FE}, but one may also see the influence of the Hubei asymmetry in, e.g., the squares corresponding to travel between Guangdong and Hubei provinces.  Finally, we randomly generate regions of unobserved ancestors from their posterior predictive distributions at every 100-thousandth MCMC iteration.  Figure \ref{fig:tree} projects the empirical posterior predictive mode of these locations onto the phylogenetic tree $\phylogeny$.

@@

In addition to these scientific questions of interest, we are interested in the performance of the first-order approximation as a surrogate gradient for HMC in such a high-dimensional setting. To isolate this performance, we fix the Bayesian bridge global-scale $\tau$ at 2.5$\times10^{-5}$.  We generate a Markov chain with 80,000 states, saving every tenth state and removing the first 1000 states as burn-in.  Despite the relatively small number of iterations, we observe large ESS that suggest satisfactory accuracy of the first-order approximation within the context of high-dimensional surrogate HMC.  The ESS for the three fixed effects regression coefficients $\theta_1$, $\theta_2$ and $\theta_3$ are 1,053, 1,343 and 498, respectively.  The median and minimum ESS for the 1,892 random effects $\epsilon_{ss'}$ are 1,514 and 1,161, respectively.

>>>>

 Let data $\y$ be the $N$-vector with elements $y_n\in \{1,\dots,S\}$, the observed states of each observed biological specimen $n \in \{1,\dots,N\}$.  At any node $v$, the likelihood takes the form \citep{ji2020gradients}
 \begin{align}
p(\y|\ttheta) = \p_v^T \q_v \, ,
 \end{align}
where $\p_v$ and $\q_v$ are the post-order and pre-order partial likelihood vectors, respectively.  The former describes the probability of the observed states for all observed specimens (i.e., leaf nodes) that descend from node $v$, conditioned on the state at node $v$. The latter describes analogous probabilities for all observed specimens not descending from node $v$.  If we suppose that node $u$ gives rise to two child nodes, $v$ and $w$, then 
\begin{align}\label{eq:recursions}
	\p_u = \P_v \p_v \circ \P_w \p_w\, ,  \quad \q_v= \P_v^T \left( \q_u \circ  \P_w \p_w\right) \, .
\end{align}
Using the chain rule and the fact that $\p_v$ does not depend on $\P_v$, one may write the likelihood's derivative with respect to a single parameter $\theta_d$ thus:
\begin{align}\label{eq:phyloGrad}
	\frac{\partial}{\partial \theta_d}  p(\y|\ttheta) 
	&=   \sum_{v=1}^{2N-2} \mbox{tr} \left( \frac{\partial (\p_v^T\q_v)}{\partial \P_v} \frac{\partial \P^T_v}{\partial \theta_d} \right)
	= \sum_{v=1}^{2N-2} \mbox{tr} \left(  (\q_u\circ \P_w\p_w) \p_v^T\bigg( \frac{\partial e^{\gamma t_v \QQ}}{\partial \theta_d}\bigg)^{\!T}\right)  \\ \nonumber
	&=\sum_{v=1}^{2N-2} \p_v^T \Biggl(\sum_{s=1}^S \sum_{s'=1}^S \frac{\partial e^{\gamma t_v \QQ}}{\partial q_{ss'}} \frac{\partial q_{ss'}}{\partial \theta_d} \Biggr)^{\!T} (\q_u\circ \P_w\p_w)  \, .
\end{align}
Whereas the recursions of \eqref{eq:recursions} facilitate fast likelihood computation, inferring $\ttheta$ using gradient-based approaches such as HMC (Section \ref{sec:surrHmc}) requires a large number of repeated evaluations of the matrix exponential derivative.  These computations become particularly onerous when one opts for a gold-standard mixed-effect model \textcolor{red}{(MAGEE)} and specifies
\begin{align}\label{eq:mixed}
\log q_{ss'} = b_{ss'} + \epsilon_{ss'} \,, \quad s\neq s'\, .
\end{align} 
Here, the fixed effects $b_{ss'}$ are elements of some non-random matrix $\B$, and the random effects $\epsilon_{ss'}$ are mutually independent \emph{a priori} and inferred as model parameters.  The dimension $D$ of $\ttheta$ in this model is $\mathcal{O}(S^2)$, so the usual $\mathcal{O}(DNS^5)$ cost of the matrix expoential derivative becomes a massive $\mathcal{O}(NS^7)$.  In this context, \textcolor{red}{MAGEE} show that the first-approximation \eqref{eq:firstOrder} achieves a considerable speedup over the exact derivative, while facilitating high-quality proposals in the context of surrogate HMC.  In the following section, we use this method to analyze the global spread of SARS-CoV-2 and show that the first-order approximation maintains its performance for an even higher-dimensional problem than previously considered.

@@

\begin{figure}[!t]
	\centering
	\includegraphics[width=\linewidth]{fixedEffects.pdf}
	\vspace{-2em}
	\caption{Matrix predictors described in \eqref{eq:linear} combine in a linear manner to form the fixed-effect matrix $\B$ featured within the mixed-effects regression model \eqref{eq:mixed}.  Air traffic proximities \emph{(\textbf{a})} are proportional the number of air passengers exchanged between airports within respective regions \citep{holbrook2021massive}.  Intracontinental proximities \emph{(\textbf{b})} take values between -1 for regions on different continents to 1 for adjacent regions.  The Hubei asymmetry \emph{(\textbf{c})} roughly characterizes the Hubei quarantine of early 2020.}\label{fig:FE}
\end{figure}

We use the phylogenetic CTMC framework to model the early spread of SARS-CoV-2---the virus responsible for the ongoing COVID-19 pandemic---based on 285 observed viral samples collected from $S=31$ regions worldwide between December 24, 2019 and March 19, 2020. These regions comprise 13 provinces within China and 18 countries without.  Understanding the manner in which viruses travel between human populations is an object of ongoing study, and phylogeographic analyses point to the central role of travel networks including those measured by airline passenger counts \citep{holbrook2021massive} or google mobility data \citep{worobey2020emergence}.  Here, we include three such predictors of travel in our CTMC model by expanding the fixed effects in regression model \eqref{eq:mixed} to take the form
\begin{align}\label{eq:linear}
	\B = \theta_1 \X_1 + \theta_2  \X_2 + \theta_3 \X_3 
\end{align} 
for each $\X$ a fixed $44\times 44$ matrix predictor and each $\theta$ a real-valued regression coefficient, where we have expanded the number of regions between which viruses may travel by including two additional Chinese provinces and 11 additional countries.  Figure \ref{fig:FE} presents the three predictors of interest: $\X_1$ contains air travel proximities between locations as measured by annualized air passenger counts between airports contained within a region \citep{holbrook2021massive};  $\X_2$ contains intracontinental proximities arising from physical distances when two regions inhabit the same continent and fixed at the minimum otherwise; finally, $\X_3$ describes the Hubei asymmetry, i.e., the non-existence of human travel out of the Hubei province in early 2020.

@@

We generate 8 million MCMC samples in this manner, saving 1 in 10,000 Markov chain states, in order to guarantee a minimum effective sample size (ESS) greater than 100.  By far, the worst-mixing parameter  is the global scale $\tau$, which obtains an ESS of 185.  The three fixed-effects regression coefficients $\theta_1$, $\theta_2$ and $\theta_3$ obtain ESS's of 721, 721 and 644, respectively.  The median and minimum ESS for the 1,892 random effects $\epsilon_{ss'}$ are 721 and 190, respectively.  We note that, after thinning and removing burn-in, the sample only consists of 721 states, so ESS of 721 implies negligible autocorrelation between samples.  The left plot of Figure \ref{fig:post} presents posterior densities for the fixed-effect coefficients from regression \eqref{eq:mixed}.  The air traffic proximity, intracontinental proximity and Hubei asymmetry coefficients have posterior means and 95\% credible intervals of 0.76 (0.50, 1.03), 0.04 (-0.04, 0.12) and 0.27 (-0.39, 0.78), respectively. While these results suggest a positive association between each predictor matrix and the infinitesimal generator matrix $\QQ$, the only predictor with a statistically significant association is air traffic proximity.  This result agrees with a previous phylogeographic analysis of the global spread of influenza \citep{holbrook2021massive}.  The right plot of Figure \ref{fig:post} presents the posterior mean for generator $\QQ$.  As one may expect, the matrix looks similar to that of the air traffic predictor matrix in Figure \ref{fig:FE}, but one may also see the influence of the Hubei asymmetry in, e.g., the squares corresponding to travel between Guangdong and Hubei provinces.  Finally, we randomly generate regions of unobserved ancestors from their posterior predictive distributions at every 100-thousandth MCMC iteration.  Figure \ref{fig:tree} projects the empirical posterior predictive mode of these locations onto the phylogenetic tree $\phylogeny$.

@@

In addition to these scientific questions of interest, we are interested in the performance of the first-order approximation as a surrogate gradient for HMC in such a high-dimensional setting. To isolate this performance, we fix the Bayesian bridge global-scale $\tau$ at 2.5$\times10^{-5}$.  We generate a Markov chain with 80,000 states, saving every tenth state and removing the first 1000 states as burn-in.  Despite the relatively small number of iterations, we observe large ESS that suggest satisfactory accuracy of the first-order approximation within the context of high-dimensional surrogate HMC.  The ESS for the three fixed-effect regression coefficients $\theta_1$, $\theta_2$ and $\theta_3$ are 1,053, 1,343 and 498, respectively.  The median and minimum ESS for the 1,892 random effects $\epsilon_{ss'}$ are 1,514 and 1,161, respectively.
